Strong lensing selection effects

### 2024-03-10 paper/seleff.tex
italics Euclid

  \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
    {
Strong lenses are a biased subset of the general population of galaxies.
}
  % aims heading (mandatory)
   {
The goal of this work is to quantify how lens galaxies and lensed sources differ from their parent distribution, namely the strong lensing bias.
} 
   % methods heading (mandatory)
   {
We first studied how the strong lensing cross-section varies as a function of lens and source properties. 
Then, we simulated strong lensing surveys with data similar to that expected for Euclid and measured the strong lensing bias in different scenarios.
We focused particularly on two quantities: the stellar population synthesis mismatch parameter, $\asps$, defined as the ratio between the true stellar mass of a galaxy and the stellar mass obtained from photometry, and the central dark matter mass at fixed stellar mass and size.
}
% results heading (mandatory)
   {
Strong lens galaxies are biased towards larger stellar masses, smaller half-mass radii and larger dark matter masses.
The amplitude of the bias depends on the intrinsic scatter in the mass-related parameters of the galaxy population and on the completeness in Einstein radius of the lens sample.
For values of the scatter that are consistent with observed scaling relations and a minimum detectable Einstein radius of $0.5''$, the strong lensing bias in $\asps$ is $10\%$, while that in the central dark matter mass is $5\%$.
The bias has little dependence on the properties of the source population: samples of galaxy-galaxy lenses and galaxy-quasar lenses that probe the same Einstein radius distribution are biased in a very similar way.
}
  % conclusions heading (optional), leave it empty if necessary 
   {
Given current uncertainties, strong lensing observations can be used directly to improve our current knowledge of the inner structure of galaxies, without the need to correct for selection effects.
Time-delay measurements of $H_0$ from lensed quasars can take advantage of prior information obtained from galaxy-galaxy lenses with similar Einstein radius.
}
   \keywords{
             Gravitational lensing: strong 
               }

@@

Strong gravitational lensing is a powerful tool for studying galaxies and cosmology.
Strong lenses have been used to probe the mass structure of massive galaxies \citep{Aug++10, ORF14, Son++15, Sha++21}, to detect substructure \citep{Veg++12, Hez++16, Nie++20}, to carry out detailed studies of magnified star-forming galaxies \citep{Jon++13}, and to measure the expansion rate of the universe with time delays \citep[see][ for a review]{T+M16}.

Strong lenses, however, are a biased subset of the general population of galaxies and background sources.
In order for a lens-source pair to be included in a strong lensing survey, the following conditions must be met:
\begin{enumerate}
\item The lens must produce multiple images of the source
\item The multiple images must be detected and resolved
\item The system must be recognised as a lens by the survey.
\end{enumerate}
Condition 1 poses constraints on the mass distribution of the lens, as well as on the geometry of the system. 
First of all, not all lenses are capable of producing multiple images, but only those whose surface mass density $\Sigma(\boldsymbol\theta)$ is larger than the critical surface mass density for lensing $\Sigma_{\mathrm{cr}}$ at at least one position $\boldsymbol\theta$ \citep{SEF92}.
This condition excludes objects with a very diffuse mass distribution from the population of lenses.
Second, objects with a higher concentration of mass allow for a larger portion of the source plane to be multiply imaged, and therefore are over-represented in lens samples. Third, at fixed lens properties, sources that are located further away along the line of sight are more likely to be lensed.

@@

Condition 3 can introduce additional biases, depending on the efficiency of the lens finding process on which the survey is based. Most commonly, lenses with a smaller image separation are more difficult to identify even if detected and resolved, because of contamination from the lens galaxy. This introduces an additional bias towards lenses with a larger or more concentrated mass.

The cumulative effect of these conditions can be summarised as follows.
The probability distribution $\prsl$ of a sample of strong lenses from a given survey with selection criterion $S$ is given by \citep{Son22}:
\begin{equation}\label{eq:one}
\prsl(\psilens,\psisource|S) \propto \prlens(\psilens)\prsource(\psisource)\psel(\psilens,\psisource|S).
\end{equation}
In the equation above, $\psilens$ is the set of parameters describing the properties of foreground galaxies that are relevant for lensing, such as their redshift and mass distribution; $\psisource$ is the set of parameters describing background sources; $\prlens$ and $\prsource$ describe the parent distribution of foreground galaxies and background sources in the absence of lensing; and $\psel$ describes the probability of selecting a lens-source system with parameters $\psilens$ and $\psisource$ given the criterion $S$ used to define a lens.
This last factor takes into account both physical effects, that is whether a lens with parameters $\psilens$ produces a strongly lensed image of a source with parameters $\psisource$, and survey selection effects, that is whether such an image can be detected and recognised as a strong lens.

The left-hand side of \Eref{eq:one} is directly accessible from strong lensing observations. If the main goal of a lensing survey is to characterise the properties of the strong lens population, then it can be accomplished by directly analysing this term. For many applications of strong lensing, however, the aim is to constrain the properties of the general galaxy or source population, $\prlens$ and $\prsource$, which are coupled in a nontrivial way via the lens selection probability $\psel$.
In order to obtain an unbiased estimate of either $\prlens$ or $\prsource$, then, it is necessary to invert \Eref{eq:one}.
In principle, this can be done with a Bayesian hierarchical formalism \citep{Son22}, but knowledge of the lens selection probability $\psel$ is required.
This factor can be written as the following product:
\begin{equation}\label{eq:two}
\psel(\psilens,\psisource|S) = \pdet(\psilens,\psisource)\pfind(\psilens,\psisource|S),
\end{equation}
where $\pdet$ is the probability of detecting a strong lensing event, and $\pfind(\psilens,\psisource|S,\mathrm{det})$ is the probability of correctly classifying it as such\footnote{\citet{Son22} implicitly assumed $\pfind\equiv1$, therefore $\pdet$ and $\psel$ can be used interchangeably in the context of that work.}.
The detection probability $\pdet$ can be obtained via simulation. The main challenge is characterising $\pfind$: in most of the existing strong lensing surveys, the process of determining whether a system is included in a strong lens sample is typically a combination of several cuts, usually involving a nontrivial visual selection step.

For the above reasons, the problem of inverting \Eref{eq:one} is a difficult one to tackle exactly.
Few studies have attempted to explicitly account for strong lensing selection effects, usually by making ad-hoc simplifying assumptions \citep{Son++15,O+A17,Son++19}.
Whether it is necessary to invert \Eref{eq:one}, however, depends on the severity of the strong lensing bias that needs to be corrected and on the accuracy requirements on the key quantities of interest.

In this paper we aim to quantify the strength of the strong lensing bias on a series of foreground galaxy and background source parameters. 
In particular, we aim to determine how strong lenses differ from the parent population of foreground galaxies and background sources in terms of a) the radial mass structure of the lenses (i.e. their stellar and dark matter mass density profiles); b) the ellipticity of the lenses; c) the size-magnitude distribution of the lensed sources.
The answer to this question depends on 1) how the lens detection probability $\pdet$ varies as a function of galaxy and source properties, 2) the efficiency of a survey at correctly classifying detected strong lenses (i.e. $\pfind$), and on 3) the shape of the galaxy and source parameters distribution $\prlens$ and $\prsource$. To understand this third point we can imagine the limiting case in which both $\prlens$ and $\prsource$ are Dirac delta functions (i.e. all lenses and sources are identical): in this limit, $\prsl$ simply reduces to the product $\prlens\prsource$ up to a scaling constant, corresponding to a case in which the lensing bias is none.

\citet{MVK09} carried out a thorough study of point 1): they quantified how the properties of a lens galaxy determine its probability of creating a lensing event with a point source.
In this work we revisited the \citet{MVK09} study, expanding it to the extended source case. We simulated individual lenses and examined how the lens detection probability varies with lens and source properties.
In addition, we addressed points 2) and 3) as well: we simulated large populations of strong lenses using empirical models, we simulated the lens detection and finding phase, and quantified the lensing bias under various scenarios.
We explored how the results change as a function of the efficiency of a lens survey at discovering small image separation lenses, and of the scatter in mass parameters at fixed light.

Finally, we considered galaxy-quasar lenses. 
The strong lensing bias of lensed quasars was studied by \citep{C+C16}, in the context of power-law lens mass models with external convergence.
Here we addressed the question of how different are galaxy-quasar lenses from galaxy-galaxy lens samples. 
This point is relevant for time-delay cosmography studies, in which measurements of the time-delay between the multiple images of a strongly lensed quasar are used to constrain the expansion rate of the Universe. Galaxy-galaxy lenses can in principle be used to help break some of the model degeneracies affecting these measurements \citep{Bir++20,B+T21}, but any difference between the two lens classes can introduce biases, if not corrected for.
With this study we aim to quantify this bias.

The structure of this work is the following.
In \Sref{sect:basics} we introduce the basics of gravitational lensing.
In \Sref{sect:indlenses} we study individual lens systems. 
In \Sref{sect:lenspop} we describe our simulations of lens surveys.
In \Sref{sect:results} we show the results of our analysis on the simulated lens survey data.
We discuss the results in \Sref{sect:discuss} and draw conclusions in \Sref{sect:concl}.

The Python code used for the simulation and analysis of the lens sample can be found in a dedicated section of a GitHub repository\footnote{\url{https://github.com/astrosonnen/strong_lensing_tools/papers/selection_effects}}.

@@

The lensing properties of an object with respect to a source depend solely on its dimensionless surface mass density distribution $\kappa(\boldsymbol\theta)$ (also referred to as the convergence). 
This is the ratio between the surface mass density and the critical surface mass density for lensing:
\begin{equation}
\kappa(\boldsymbol\theta) = \dfrac{\Sigma(\boldsymbol\theta)}{\Sigma_{\mathrm{cr}}}.
\end{equation}
The latter quantity is defined as
\begin{equation}
\Sigma_{\mathrm{cr}} = \dfrac{c^2D_\mathrm{s}}{4\pi G D_\mathrm{d} D_{\mathrm{ds}}},
\end{equation}
where $c$ is the speed of light, $G$ the gravitational constant, and $D_\mathrm{d}$, $D_\mathrm{s}$, and $D_{\mathrm{ds}}$ are the angular diameter distances between the observer and the lens, the observer and the source, and the lens and the source, respectively.

@@

In the special case of an axisymmetric lens we can simplify the notation by considering a single coordinate axis with origin at the centre of the lens. We label the components of the image position, source position and deflection angle along this axis as $\theta$, $\beta$ and $\alpha$, respectively.
The lens equation for an axisymmetric lens then becomes
\begin{equation}\label{eq:1dlenseq}
\beta = \theta - \alpha(\theta),
\end{equation}
and the expression for the deflection angle reduces to
\begin{equation}
\alpha(\theta) = \frac{2}{\theta}\int_0^{\theta} d\theta' \kappa(\theta') \theta'.
\end{equation}
This can also be expressed in terms of the total projected mass enclosed within a circle with angular radius equal to $\theta$:
\begin{equation}
\alpha(\theta) = \frac{1}{\pi\theta}\dfrac{M(<\theta)}{\Sigma_{\mathrm{cr}}D_{\mathrm{d}}^2}.
\end{equation}

A very important quantity for describing the strength of a strong lens is the Einstein radius, $\tein$. For an axisymmetric lens, this is the radius corresponding to the solution of \Eref{eq:1dlenseq} for a source placed at the same angular position as the lens centre ($\beta=0$).
The circle with radius equal to $\tein$ is known as the tangential critical curve. Images that form there have infinite magnification in the tangential direction.
%
It can be shown that $\tein$ satisfies the following condition,
\begin{equation}\label{eq:criteq}
\bar{\kappa}(<\tein) = 1,
\end{equation}
where $\bar{\kappa}(<\theta)$ is the average surface mass density within a radius equal to $\theta$:
\begin{equation}
\bar{\kappa}(<\theta) \equiv \frac{2}{\theta^2}\int_0^{\theta}d\theta' \kappa(\theta')\theta'.
\end{equation}

Axisymmetric lenses of the kind considered in this work 
typically produce either one or three images of a point source.
\Fref{fig:scheme} helps to visualise this property.
Plotted in \Fref{fig:scheme} is the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, for a few lens models.
According to \Eref{eq:1dlenseq}, images form at the locations where this quantity equals the position of the source. 
Therefore, given a source position $\beta$, the number of images and their location can be determined by drawing a horizontal line at the value $\beta$ on the vertical axis, and finding the points where this line intersects the $\theta - \alpha(\theta)$ curve.

For small values of $\beta$ (for sources close to the lens centre), the number of images that are produced is three: the source is strongly lensed. For large values of $\beta$, instead, only one image is formed.
The value of $\beta$ where the transition occurs is known as the radial caustic, which is marked by the horizontal dotted line in \Fref{fig:scheme}. 
As can be seen from \Fref{fig:scheme}, a source at this location is mapped to the stationary point of the $\theta - \alpha(\theta)$ curve. 
That location on the image plane is known as the radial critical curve. 
Images that form there have a formally infinite magnification in the radial direction\footnote{The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the radial magnification. This can be understood by taking the derivative of \Eref{eq:1dlenseq} with respect to $\theta$.}.

\begin{figure}
\includegraphics[width=\columnwidth]{composite_fixedap_scheme.eps}
\caption{
The lens equation of axisymmetric lenses. Solid lines: right-hand side of \Eref{eq:1dlenseq} for five lenses with different density profiles. 
Given a source at position $\beta$, its lensed images form at the values of $\theta$ where the solid line intersects a horizontal line located at $\beta$ on the vertical axis. 
Dotted lines: positions of the radial caustics. Sources located within the radial caustic of a given lens produce three images.
The lens models used in this simulations consist of a stellar component and a dark matter halo, as described in section \ref{ssec:profile}. Their density profile is plotted in \Fref{fig:kappa}.
\label{fig:scheme}
}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{composite_caustics.eps}
\caption{
Caustics of lenses with fixed radial structure and different ellipticity.
Source-plane angular coordinates are in units of the Einstein radius.
The outer curve is the radial caustic, while the inner diamond is the tangential caustic. Point sources located outside the caustic are not strongly lensed. Sources that lie in the region enclosed between the two caustics produce three images, while sources inside the tangential caustic produce five images.
The lens model adopted for this experiment consists of a stellar component and a dark matter halo, as described in \ref{ssec:profile}. The two components have the same ellipticity.
\label{fig:ellcaust}
}
\end{figure}

\subsection{Lensing event definition: point sources}\label{ssec:lensdefpoint}

@@

\subsection{Lensing event definition: extended sources}\label{ssec:lensdefext}

Defining a strong lensing event in the case of an extended source is less straightforward.
In principle, we should require parts of the source to be multiply imaged.
In practice, it is not always easy to determine whether a lensed image contains multiple images or not.
This is because, when the source is extended, some of the images can be blended together.
In real strong lensing surveys, it is common to find lens candidates in which the lensed source consists of only a single arc. In those cases it is difficult to establish whether the arc is a set of blending images or not, and the decision of including such systems in a strong lens sample is often arbitrary.
Here we adopt the following working definition: {\em an extended source is strongly lensed if the number of surface brightness peaks that are detected is larger than its intrinsic number of peaks in the absence of lensing}.

@@

We point out that, although our lens definition relies on peak detections, it is not necessary for the peak of the source surface brightness to lie within the caustics in order for the source to be strongly lensed.
This can be observed in the fourth example of \Fref{fig:lensdef}: although the source centroid lies outside of the caustic (as shown in the first column), the outskirts of the source overlap with the lens, and therefore multiple images are produced.

\begin{figure}
\includegraphics[width=\columnwidth]{example_detections.eps}
\caption{
Criterion used to classify lensed images of extended sources. Six examples.
First column: caustics (red curves) and source position (blue circle). The radius of the circle corresponds to the radius at which the surface brightness is equal to $2\sigma$ the sky background rms fluctuation for a single pixel. In other words, the blue circle delimits the area of the source that can be detected.
Second column: mock image of the lensed source, with added noise.
Third column: footprint of the source. The purple footprint is obtained with a $2\sigma$ detection criterion. The coral region is the largest footprint with the highest number of detected images.
\label{fig:lensdef}
}
\end{figure}

We can identify three different regimes in strong lensing of extended sources, depending on the relative size of the source and of the lens caustics.
In the limit of a very small source, the image configurations that are produced are qualitatively similar to those that can be obtained in the point-source case.
When the source and caustic size are comparable, the multiple images tend to be blended into arcs or rings. 
Most, if not all, of the galaxy-scale strong lenses known belong to these two categories.
However, there is a third regime in strong lensing, corresponding to the case in which the source size is much bigger than the caustics, such as in the fifth example of \Fref{fig:lensdef}.
In this regime, the overall size and total flux of the source are roughly preserved, and the lens produces only a relatively minor perturbation on a localised region of the image.
Strong lenses of this kind can be difficult to detect, especially if the region of the image subject to strong lensing overlaps with the light from the foreground galaxy. 
The ultimate limiting factor to the detection of strong lenses in the large source size regime, however, is the ability to spatially resolve the multiple images. This limit is set by the size of the point spread function (PSF).

@@

We fix the S\'{e}rsic index of the lenses to $n=4$, corresponding to a de Vaucouleurs model.
Although early-type galaxies are often described with a free S\'{e}rsic index, a de Vaucouleurs profile is able to reproduce their surface brightness distribution to a few percent in the radial range $1\rm{kpc} < R < 30\rm{kpc}$ \citep[see e.g.][]{Son20}, which is the region that is most relevant for strong lensing.
Finally, we assume that the light distribution of the stellar component follows its mass distribution exactly. That is, we do not allow for the presence of gradients in the stellar mass-to-light ratio. In section \ref{ssec:limitations} we discuss qualitatively what the implications of relaxing this assumption would be.

>>>>

  \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
    {
Strong lenses are a biased subset of the general population of galaxies.
}
  % aims heading (mandatory)
   {
The goal of this work is to quantify how lens galaxies and lensed sources differ from their parent distribution, namely the strong lensing bias.
} 
   % methods heading (mandatory)
   {
We first studied how the strong lensing cross-section varies as a function of lens and source properties. 
Then, we simulated strong lensing surveys with data similar to that expected for \textit{Euclid} and measured the strong lensing bias in different scenarios.
We focused particularly on two quantities: the stellar population synthesis mismatch parameter, $\asps$, defined as the ratio between the true stellar mass of a galaxy and the stellar mass obtained from photometry, and the central dark matter mass at fixed stellar mass and size.
}
% results heading (mandatory)
   {
Strong lens galaxies are biased towards higher stellar masses, smaller half-mass radii, and higher dark matter masses.
The amplitude of the bias depends on the intrinsic scatter in the mass-related parameters of the galaxy population and on the completeness in Einstein radius of the lens sample.
For values of the scatter that are consistent with observed scaling relations and a minimum detectable Einstein radius of $0.5''$, the strong lensing bias in $\asps$ is $10\%$, while that in the central dark matter mass is $5\%$.
The bias has little dependence on the properties of the source population: samples of galaxy-galaxy lenses and galaxy-quasar lenses that probe the same Einstein radius distribution are biased in a very similar way.
}
  % conclusions heading (optional), leave it empty if necessary 
   {
Given current uncertainties, strong lensing observations can be used directly to improve our current knowledge of the inner structure of galaxies, without the need to correct for selection effects. Time-delay measurements of $H_0$ from lensed quasars can take advantage of prior information obtained from galaxy-galaxy lenses with similar Einstein radii.
}
   \keywords{
             Gravitational lensing: strong 
               }

@@

Strong gravitational lensing is a powerful tool for studying galaxies and cosmology.
Strong lenses have been used to probe the mass structure of massive galaxies \citep{Aug++10, ORF14, Son++15, Sha++21}, to detect substructure \citep{Veg++12, Hez++16, Nie++20}, to carry out detailed studies of magnified star-forming galaxies \citep{Jon++13}, and to measure the expansion rate of the Universe with time delays \citep[see][ for a review]{T+M16}.

Strong lenses, however, are a biased subset of the general population of galaxies and background sources.
In order for a lens-source pair to be included in a strong lensing survey, the following conditions must be met: (1) the lens must produce multiple images of the source, (2) the multiple images must be detected and resolved, and (3) the system must be recognised as a lens by the survey.

Condition 1 poses constraints on the mass distribution of the lens, as well as on the geometry of the system. 
First of all, not all lenses are capable of producing multiple images, only those whose surface mass density, $\Sigma(\boldsymbol\theta)$, is higher than the critical surface mass density for lensing, $\Sigma_{\mathrm{cr}}$, at at least one position, $\boldsymbol\theta$ \citep{SEF92}.
This condition excludes objects with a very diffuse mass distribution from the population of lenses.
Second, objects with a higher concentration of mass allow for a larger portion of the source plane to be multiply imaged and are therefore over-represented in lens samples. Third, with fixed lens properties, sources that are located farther away along the line of sight are more likely to be lensed.

@@

Condition 3 can introduce additional biases, depending on the efficiency of the lens finding process on which the survey is based. Most commonly, lenses with a smaller image separation are more difficult to identify even if detected and resolved, because of contamination from the lens galaxy. This introduces an additional bias towards lenses with a higher or more concentrated mass.

The cumulative effect of these conditions can be summarised as follows.
The probability distribution, $\prsl$, of a sample of strong lenses from a given survey with selection criterion $S$ is given by \citep{Son22}
\begin{equation}\label{eq:one}
\prsl(\psilens,\psisource|S) \propto \prlens(\psilens)\prsource(\psisource)\psel(\psilens,\psisource|S).
\end{equation}
In this equation, $\psilens$ is the set of parameters describing the properties of foreground galaxies that are relevant for lensing, such as their redshift and mass distribution; $\psisource$ is the set of parameters describing background sources; $\prlens$ and $\prsource$ describe the parent distribution of foreground galaxies and background sources in the absence of lensing; and $\psel$ describes the probability of selecting a lens-source system with parameters $\psilens$ and $\psisource$ given the criterion, $S$, used to define a lens.
This last factor takes both physical and survey selection effects into account, that is, whether a lens with parameters $\psilens$ produces a strongly lensed image of a source with parameters $\psisource$, and whether such an image can be detected and recognised as a strong lens.

The information needed for the left-hand side of Eq. \ref{eq:one} is directly accessible from strong lensing observations. If the main goal of a lensing survey is to characterise the properties of the strong lens population, then it can be accomplished by directly analysing this term. For many applications of strong lensing, however, the aim is to constrain the properties of the general galaxy or source population, $\prlens$ and $\prsource$, which are coupled in a non-trivial way via the lens selection probability, $\psel$.
In order to obtain an unbiased estimate of either $\prlens$ or $\prsource$, it is necessary to invert Eq. \ref{eq:one}.
In principle, this can be done with a Bayesian hierarchical formalism \citep{Son22}, but knowledge of the lens selection probability, $\psel$, is required.
This factor can be written as the following product:
\begin{equation}\label{eq:two}
\psel(\psilens,\psisource|S) = \pdet(\psilens,\psisource)\pfind(\psilens,\psisource|S),
\end{equation}
where $\pdet$ is the probability of detecting a strong lensing event, and $\pfind(\psilens,\psisource|S,\mathrm{det})$ is the probability of correctly classifying it as such\footnote{\citet{Son22} implicitly assumed $\pfind\equiv1$; therefore, $\pdet$ and $\psel$ can be used interchangeably in the context of that work.}.
The detection probability, $\pdet$, can be obtained via simulations. The main challenge is characterising $\pfind$: in most of the existing strong lensing surveys, the process of determining whether a system is included in a strong lens sample is typically a combination of several cuts, usually involving a non-trivial visual selection step.

For the above reasons, the problem of inverting Eq. \ref{eq:one} is a difficult one to tackle exactly.
A few studies have attempted to explicitly account for strong lensing selection effects, usually by making ad hoc simplifying assumptions \citep{Son++15,O+A17,Son++19}.
Whether it is necessary to invert Eq. \ref{eq:one}, however, depends on the severity of the strong lensing bias that needs to be corrected and on the accuracy requirements for the key quantities of interest.

In this paper we aim to quantify the strength of the strong lensing bias on a series of foreground galaxy and background source parameters. 
In particular, we aim to determine how strong lenses differ from the parent population of foreground galaxies and background sources in terms of (a) the radial mass structure of the lenses (i.e. their stellar and dark matter mass density profiles), (b) the ellipticity of the lenses, and (c) the size-magnitude distribution of the lensed sources.
This determination depends on (1) how the lens detection probability, $\pdet$, varies as a function of galaxy and source properties, (2) the efficiency of a survey in correctly classifying detected strong lenses (i.e. $\pfind$), and (3) the shape of the galaxy and source parameter distribution, $\prlens$ and $\prsource$. To understand this third point we can imagine the limiting case in which both $\prlens$ and $\prsource$ are Dirac delta functions (i.e. all lenses and sources are identical): in this limit, $\prsl$ simply reduces to the product $\prlens\prsource$ up to a scaling constant; this corresponds to a case in which there is no lensing bias.

\citet{MVK09} carried out a thorough study of point (1): they quantified how the properties of a lens galaxy determine its probability of creating a lensing event with a point source.
In this work we revisit the \citet{MVK09} study, expanding it to the extended source case. We simulated individual lenses and examined how the lens detection probability varies with lens and source properties.
In addition, we address points (2) and (3): we simulated large populations of strong lenses using empirical models, we simulated the lens detection and finding phase, and we quantified the lensing bias under various scenarios.
We explored how the results change as a function of the efficiency of a lens survey in discovering small image separation lenses, and as a function of the scatter in mass parameters at fixed light.

Finally, we considered galaxy-quasar lenses. 
The strong lensing bias of lensed quasars was studied by \cite{C+C16} in the context of power-law lens mass models with external convergence.
Here we address the question of how different galaxy-quasar lenses are from galaxy-galaxy lens samples. 
This point is relevant for time-delay cosmography studies, in which measurements of the time delay between the multiple images of a strongly lensed quasar are used to constrain the expansion rate of the Universe. Galaxy-galaxy lenses can in principle be used to help break some of the model degeneracies affecting these measurements \citep{Bir++20,B+T21}, but any difference between the two lens classes can introduce biases if not corrected for.
With this study we aim to quantify this bias.

The structure of this paper is as follows.
In Sect. \ref{sect:basics} we introduce the basics of gravitational lensing.
In Sect. \ref{sect:indlenses} we study individual lens systems. 
In Sect. \ref{sect:lenspop} we describe our simulations of lens surveys.
In Sect. \ref{sect:results} we show the results of our analysis of the simulated lens survey data.
We discuss the results in Sect. \ref{sect:discuss} and draw conclusions in Sect. \ref{sect:concl}.

The Python code used for the simulation and analysis of the lens sample can be found in a dedicated section of a GitHub repository\footnote{\url{https://github.com/astrosonnen/strong_lensing_tools/tree/main/papers/selection_effects}}.

@@

The lensing properties of an object with respect to a source depend solely on its dimensionless surface mass density distribution, $\kappa(\boldsymbol\theta)$ (also referred to as the convergence). 
This is the ratio between the surface mass density and the critical surface mass density for lensing:
\begin{equation}
\kappa(\boldsymbol\theta) = \dfrac{\Sigma(\boldsymbol\theta)}{\Sigma_{\mathrm{cr}}}.
\end{equation}
The latter quantity is defined as
\begin{equation}
\Sigma_{\mathrm{cr}} = \dfrac{c^2D_\mathrm{s}}{4\pi G D_\mathrm{d} D_{\mathrm{ds}}},
\end{equation}
where $c$ is the speed of light, $G$ the gravitational constant, and $D_\mathrm{d}$, $D_\mathrm{s}$, and $D_{\mathrm{ds}}$ are the angular diameter distances between the observer and the lens, the observer and the source, and the lens and the source, respectively.

@@

In the special case of an axisymmetric lens we can simplify the notation by considering a single coordinate axis with origin at the centre of the lens. We label the components of the image position, source position and deflection angle along this axis as $\theta$, $\beta$, and $\alpha$, respectively.
The lens equation for an axisymmetric lens then becomes
\begin{equation}\label{eq:1dlenseq}
\beta = \theta - \alpha(\theta),
\end{equation}
and the expression for the deflection angle reduces to
\begin{equation}
\alpha(\theta) = \frac{2}{\theta}\int_0^{\theta} d\theta' \kappa(\theta') \theta'.
\end{equation}
This can also be expressed in terms of the total projected mass enclosed within a circle with angular radius equal to $\theta$:
\begin{equation}
\alpha(\theta) = \frac{1}{\pi\theta}\dfrac{M(<\theta)}{\Sigma_{\mathrm{cr}}D_{\mathrm{d}}^2}.
\end{equation}

A very important quantity for describing the strength of a strong lens is the Einstein radius, $\tein$. For an axisymmetric lens, this is the radius corresponding to the solution of Eq. \ref{eq:1dlenseq} for a source placed at the same angular position as the lens centre ($\beta=0$).
The circle with radius equal to $\tein$ is known as the tangential critical curve. Images that form there have infinite magnification in the tangential direction.
%
It can be shown that $\tein$ satisfies the following condition,
\begin{equation}\label{eq:criteq}
\bar{\kappa}(<\tein) = 1,
\end{equation}
where $\bar{\kappa}(<\theta)$ is the average surface mass density within a radius equal to $\theta$:
\begin{equation}
\bar{\kappa}(<\theta) \equiv \frac{2}{\theta^2}\int_0^{\theta}d\theta' \kappa(\theta')\theta'.
\end{equation}

Axisymmetric lenses of the kind considered in this work 
typically produce either one or three images of a point source.
\Fref{fig:scheme} helps visualise this property.
Plotted in \Fref{fig:scheme} is the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, for a few lens models.
According to Eq. \ref{eq:1dlenseq}, images form at the locations where this quantity equals the position of the source. 
Therefore, given a source position $\beta$, the number of images and their location can be determined by drawing a horizontal line at the value $\beta$ on the vertical axis, and finding the points where this line intersects the $\theta - \alpha(\theta)$ curve.

For small values of $\beta$ (for sources close to the lens centre), the number of images that are produced is three: the source is strongly lensed. For large values of $\beta$, instead, only one image is formed.
The value of $\beta$ where the transition occurs is known as the radial caustic, which is marked by the horizontal dotted line in \Fref{fig:scheme}. 
As can be seen from \Fref{fig:scheme}, a source at this location is mapped to the stationary point of the $\theta - \alpha(\theta)$ curve. 
That location on the image plane is known as the radial critical curve. 
Images that form there have a formally infinite magnification in the radial direction\footnote{The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the radial magnification. This can be understood by taking the derivative of Eq. \ref{eq:1dlenseq} with respect to $\theta$.}.

\begin{figure}
\includegraphics[width=\columnwidth]{composite_fixedap_scheme.eps}
\caption{
Lens equation of axisymmetric lenses. Solid lines represent the right-hand side of Eq. \ref{eq:1dlenseq} for five lenses with different density profiles. 
Given a source at position $\beta$, its lensed images form at the values of $\theta$ where the solid line intersects a horizontal line located at $\beta$ on the vertical axis. 
Dotted lines represent positions of the radial caustics. Sources located within the radial caustic of a given lens produce three images.
The lens models used in this simulation consist of a stellar component and a dark matter halo, as described in Sect. \ref{ssec:profile}. Their density profile is plotted in Fig. \ref{fig:kappa}.
\label{fig:scheme}
}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\columnwidth]{composite_caustics.eps}
\caption{
Caustics of lenses with fixed radial structure and different ellipticity.
Source-plane angular coordinates are in units of the Einstein radius.
The outer curve is the radial caustic, while the inner diamond is the tangential caustic. Point sources located outside the caustic are not strongly lensed. Sources that lie in the region enclosed by the two caustics produce three images, while sources inside the tangential caustic produce five images.
The lens model adopted for this experiment consists of a stellar component and a dark matter halo, as described in Sect. \ref{ssec:profile}. The two components have the same ellipticity.
\label{fig:ellcaust}
}
\end{figure}

\subsection{Lensing event definition: Point sources}\label{ssec:lensdefpoint}

@@

\subsection{Lensing event definition: Extended sources}\label{ssec:lensdefext}

Defining a strong lensing event in the case of an extended source is less straightforward.
In principle, we should require parts of the source to be multiply imaged.
In practice, it is not always easy to determine whether a lensed image contains multiple images or not.
This is because, when the source is extended, some of the images can be blended together.
In real strong lensing surveys, it is common to find lens candidates in which the lensed source consists of only a single arc. In those cases it is difficult to establish whether the arc is a set of blending images or not, and the decision of including such systems in a strong lens sample is often arbitrary.
Here we adopt the following working definition: an extended source is strongly lensed if the number of surface brightness peaks that are detected is larger than its intrinsic number of peaks in the absence of lensing.

@@

We point out that, although our lens definition relies on peak detections, it is not necessary for the peak of the source surface brightness to lie within the caustics in order for the source to be strongly lensed.
This can be observed in the fourth example of Fig. \ref{fig:lensdef}: although the source centroid lies outside of the caustic (as shown in the first column), the outskirts of the source overlap with the lens, and therefore multiple images are produced.

\begin{figure}
\includegraphics[width=\columnwidth]{example_detections.eps}
\caption{
Criterion used to classify lensed images of extended sources. Six examples are shown.
First column: Caustics (red curves) and source position (blue circle). The radius of the circle corresponds to the radius at which the surface brightness is equal to $2\sigma$ the sky background rms fluctuation for a single pixel. In other words, the blue circle delimits the area of the source that can be detected.
Second column: Mock image of the lensed source, with added noise.
Third column: Footprint of the source. The purple footprint is obtained with a $2\sigma$ detection criterion. The coral region is the largest footprint with the highest number of detected images.
\label{fig:lensdef}
}
\end{figure}

We can identify three different regimes in strong lensing of extended sources, depending on the relative size of the source and of the lens caustics.
In the limit of a very small source, the image configurations that are produced are qualitatively similar to those that can be obtained in the point-source case.
When the source and caustic size are comparable, the multiple images tend to be blended into arcs or rings. 
Most, if not all, of the galaxy-scale strong lenses known belong to these two categories.
However, there is a third regime in strong lensing, corresponding to the case in which the source size is much bigger than the caustics, such as in the fifth example of Fig. \ref{fig:lensdef}.
In this regime, the overall size and total flux of the source are roughly preserved, and the lens produces only a relatively minor perturbation on a localised region of the image.
Strong lenses of this kind can be difficult to detect, especially if the region of the image subject to strong lensing overlaps with the light from the foreground galaxy. 
The ultimate limiting factor to the detection of strong lenses in the large source size regime, however, is the ability to spatially resolve the multiple images. This limit is set by the size of the point spread function (PSF).

@@

We fix the S\'{e}rsic index of the lenses to $n=4$, corresponding to a de Vaucouleurs model.
Although early-type galaxies are often described with a free S\'{e}rsic index, a de Vaucouleurs profile is able to reproduce their surface brightness distribution to a few percent in the radial range $1\rm{kpc} < R < 30\rm{kpc}$ \citep[see e.g.][]{Son20}, which is the region that is most relevant for strong lensing.
Finally, we assume that the light distribution of the stellar component follows its mass distribution exactly. That is, we do not allow for the presence of gradients in the stellar mass-to-light ratio. In Sect. \ref{ssec:limitations} we discuss qualitatively what the implications of relaxing this assumption would be.
